---
title: "Music Recommendation with Association Rule Mining"
author: "Stefani DW Yates"
date: "2025-03-9"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cerulean
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

# Introduction

Online radio platforms track user listening behavior to suggest music.
We apply association rule mining using Apriori to uncover patterns in a Last.fm dataset.

```{r load_libraries}
library(arules)
library(tidyverse)
library(Matrix)
library(arulesViz)
library(igraph)
library(viridis)
library(knitr)
library(ggplot2)
```

## Load and Explore Dataset
```{r EDA}
file_path_main <- "C:/Users/stefa/OneDrive - Careered - CTU/2025/CS871/Week 9/lastfm-dataset-360K/lastfm-dataset-360K/usersha1-artmbid-artname-plays.tsv"
file_path_profile <- "C:/Users/stefa/OneDrive - Careered - CTU/2025/CS871/Week 9/lastfm-dataset-360K/lastfm-dataset-360K/usersha1-profile.tsv"

data_main <- read.delim(file_path_main, header = FALSE, sep = "\t", quote = "", fill = TRUE, stringsAsFactors = FALSE)
data_profile <- read.delim(file_path_profile, header = FALSE, sep = "\t", quote = "", fill = TRUE, stringsAsFactors = FALSE)

colnames(data_main) <- c("user_id", "artist_id", "artist_name", "plays")
colnames(data_profile) <- c("user_id", "gender", "age", "country", "signup_date")
```

### Summary Statistics
```{r summary_stats}
str(data_main)
str(data_profile)
```
```{r missing_values}
sum(is.na(data_main))
sum(is.na(data_profile))
```

### Top 10 Artists
```{r top10, fig.align="center", fig.width=8}
top_artists <- data_main %>%
  group_by(artist_name) %>%
  summarise(total_plays = sum(plays)) %>%
  arrange(desc(total_plays)) %>%
  slice_head(n = 10)

kable(top_artists)
```

## Data Preprocessing
### All Artists
```{r 104}
top_104_artists <- data_main %>%
  group_by(artist_name) %>%
  summarise(total_plays = sum(plays)) %>%
  arrange(desc(total_plays)) %>%
  slice_head(n = 104) %>%
  pull(artist_name)

filtered_data <- data_main %>%
  filter(artist_name %in% top_104_artists) %>%
  mutate(value = 1) %>%
  select(user_id, artist_name, value)
```

### Matrix Conversion
```{r artist_matrix}
user_artist_matrix <- filtered_data %>%
  pivot_wider(names_from = artist_name, values_from = value, values_fill = list(value = 0)) %>%
  select(-user_id) %>%
  as.matrix()

# Convert to transactions
user_artist_matrix[user_artist_matrix > 1] <- 1  
trans <- as(user_artist_matrix, "transactions")

summary(trans)
```

## Rule Mining
### Applying *a priori* Algorithm
```{r rules}
rules <- apriori(trans, parameter = list(supp = 0.005, conf = 0.5, minlen = 2))

# View top 10 rules
inspect(sort(rules, by = "lift")[1:10])
```

## Visualizations
### Top 50 Rules Cluster Graph
```{r cluster50, fig.align="center", fig.width=10, fig.height=8}
subrules_50 <- head(sort(rules, by = "lift"), 50)
rules_graph <- plot(subrules_50, method = "graph", engine = "igraph", plot = FALSE)

plot(rules_graph,
     layout = layout_with_fr(rules_graph),
     vertex.size = 8,
     vertex.label.cex = 1.2,
     vertex.label.color = "black",
     edge.arrow.size = 0.5)
```

## Scatter Plot of All Rules
```{r scatter, fig.align="center", fig.width=12, fig.height=8}
rules_large <- apriori(trans, parameter = list(supp = 0.003, conf = 0.6, minlen = 2, maxlen = 5))
scatterplot_rules <- subset(rules_large, lift >= 3 & confidence >= 0.7)

plot(scatterplot_rules, method = "scatterplot", measure = c("support", "lift"), shading = "confidence")
```

## Matrix Plot
```{r matrix, fig.align="center", fig.width=12, fig.height=8}
plot(subrules_50, method = "matrix", measure = "lift", 
     control = list(reorder = "support/confidence", col = viridis::mako(2.5)))
```

## Grouped Plot
```{r grouped, fig.align="center", fig.width=12, fig.height=8}
plot(subrules_50, method = "grouped", control = list(col = hcl.colors(5, "Blues")))
```

## Summary

✔ Successfully processed Last.fm dataset (360K users, 17M records).

✔ Built a user-artist matrix for association rule mining.

✔ Discovered artist relationships using Apriori.

✔ Visualized results with cluster graphs, scatter plots, and matrix views.

----------------------------------------------------------------------
### References

Celma, O. (2009). *Music recommendation datasets for research.* Universitat Pompeu Fabra. http://www.dtic.upf.edu/~ocelma/MusicRecommendationDataset/lastfm-360K.html

Csardi, G., & Nepusz, T. (2023). *The igraph software package for complex network research*. InterJournal, Complex Systems, [Version info]. https://igraph.org

Hahsler, M., Gruen, B., & Hornik, K. (2023). *arules: Mining Association Rules and Frequent Itemsets*. R package version [Version info]. https://CRAN.R-project.org/package=arules